Perturbing SU($n$)-symmetric interactions

### 2020-02-04 collective_sun/sun_perturbation.tex
modified the definition of diagrams to simply the rules used to decompose them

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  We then define
the restriction of $k\in\C_N\p{g}$ to values associated with $w_j$ and
$O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} { M_j \choose g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  { M \choose a_1, a_2, \cdots, a_m }
  \equiv \f{M!}{a_1!a_2!\cdots a_m!},
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices according to $g_S$ for all
$j\in S$, and the second product in \eqref{eq:sym_prod_symmetry}
accounts for the number of ways to permute each set of shared indices
on every tensor.  Altogether, the symmetry factors $\S\p{g}$ cancel
out with the $\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

@@

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the permutationally symmetric manifold $\M_0$ by the perturbation
$\V_1$ in \eqref{eq:perturbations} requires us to simplify a product
of the form

\begin{align}
  \P_0 \sp{\sum_{p,q} v_p w_q X_p Y_q} \P_0
  \EQFS \diagram{single_body_0} X_1 Y_2
  + \diagram{single_body_1} X_1 Y_1,
  \label{eq:PXYP_start}
\end{align}
where $p,q$ index individual spins; $X,Y$ are single-spin operators;
and the coefficients $v_p,w_q$ satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta\ne0$, which implies that they are mean-zero.
Writing out
\begin{align}
  \diagram{single_body_1}
  \equiv \sum_p v_p w_p
  = \v v\c\v w,
\end{align}
\begin{align}
  \diagram{single_body_0}
  \equiv \sum_{p\ne q} v_p w_q
  = \sum_{p,q} v_p w_q - \sum_{p=q} v_p w_q
  = - \v v \c\v w,
\end{align}
we thus have
\begin{align}
  \sum_{p,q} v_p w_q X_p Y_q
  \EQFS - \v v\c\v w \p{X_1 Y_2 - X_1 Y_1},
\end{align}
This result is also simple to write in terms of collective operators,
as
\begin{align}
  \col{X Y} \EQFS N X_1 Y_1,
  &&
  \col{Z} \equiv \sum_p Z_p,
\end{align}
\begin{align}
  \col{X}\,\col{Y}
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_1 Y_2,
\end{align}
so
\begin{align}
  \sum_{p,q} v_p w_q X_p Y_q
  \EQFS - \f{\v v \c\v w}{N\p{N-1}}
  \p{\col{X}\,\col{Y} - N \col{XY}}.
\end{align}

@@

Calculating the second-order effective Hamiltonian $H_2^{(2)}$ induced
on the permutationally symmetric manifold $\M_0$ by the perturbation
$\V_2$ in \eqref{eq:perturbations} requires us to simplify
\begin{multline}
  \P_0 \sp{\sum_{k<\ell} \sum_{p<q}
    v_{k\ell} w_{pq} O_{k\ell} Q_{pq}} \P_0 \\
  \EQFS \diagram{two_body_0} O_{1,2} Q_{3,4}
  + \diagram{two_body_1} O_{1,2} Q_{1,3}
  + \diagram{two_body_2} O_{1,2} Q_{1,2},
  \label{eq:POQP_start}
\end{multline}
where $k,\ell,p,q$ index individual spins; $O$ and $Q$ are
permutationally-symmetric two-spin operators; and the coefficients
$v_{k\ell},w_{pq}$ satisfy \eqref{eq:cond_2} with eigenvalue
$\Delta\ne0$, which implies that they are mean-zero.  Defining
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
\end{align}
for $m\in\set{v,w}$, we can write out explicitly
\begin{align}
  \diagram{two_body_2}
  \equiv \sum_{p<q} v_{pq} w_{pq} = \v{\m v} \c\v{\m w},
\end{align}
and, using the fact that $v_{pp}=w_{pp}=0$,
\begin{align}
  \diagram{two_body_1}
  \equiv \sum_{\substack{k,p,q\\k\ne p\ne q\ne k}} v_{kp} w_{kq}
  = \sum_{k,p} v_{kp} \sum_{q\ne p} w_{kq}
  = \sum_{k,p} v_{kp} \p{w_{k} - w_{kp}}
  = \v v \c \v w - 2 \v{\m v} \c \v{\m w}.
\end{align}
Using additionally the fact that $v_{pq},w_{pq}$ are mean-zero, with
some more work we find that
\begin{align}
  \diagram{two_body_0}
  \equiv \sum_{k<\ell} \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}}
  v_{k\ell} w_{pq} = -\v v \c \v w + \v{\m v} \c \v{\m w}.
\end{align}
Altogether, we thus have that
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} O_{k\ell} Q_{pq} \\
  \EQFS - \p{\v v \c \v w - \v{\m v} \c \v{\m w}} O_{1,2} Q_{3,4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} O_{1,2} Q_{1,3}
  + \v{\m v} \c \v{\m w}\, O_{1,2} Q_{1,2}.
  \label{eq:POQP}
\end{multline}

>>>>

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
  \label{eq:index_values}
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  In words,
$\C_N\p{g}$ consists of all the ways to choose, for each subset $S$ of
the tensors $\set{w_j}$, a set $k_S$ of $g_S$ distinct spins, with the
restriction that all $k_S$ are orthogonal: if $z\in k_S$, then
$z\notin k_R$ for $R\ne S$.  Having defined the set $\C_N\p{g}$ of
index values, we define the restriction of $k\in\C_N\p{g}$ to values
associated with $w_j$ and $O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
  \label{eq:diagram_factors}
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} { M_j \choose g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  { M \choose a_1, a_2, \cdots, a_m }
  \equiv \f{M!}{a_1!a_2!\cdots a_m!}
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices according to $g_S$ for all
$j\in S$, and the second product in \eqref{eq:sym_prod_symmetry}
accounts for the number of ways to permute each set of shared indices
on every tensor.  Altogether, the symmetry factors $\S\p{g}$ cancel
out with the $\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQFS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Formalizing diagrams and reducing computational costs}
\label{sec:diagrams}

Diagrams such as \eqref{eq:venn_diagram} will play a major role in the
calculations in this section.  Here, we first define diagrams as
formal objects that we can use in mathematical expressions.
Formalizing diagrams will enable us to perform necessary calculations
that are otherwise intractable.  We then discuss rules for decomposing
these diagrams in order to reduce the computational cost of their
numerical evaluation.

The simplest diagrams have $g_S$ dots (``$\bullet$'') at the
intersection of the circles $j\in S$, and are proportional to the
associated scalar $w\p{g}$ defined in \eqref{eq:diagram_factors}, e.g.
\begin{align}
  \diagram{example_triple}
  \equiv \sp{\prod_{S\in\PP\p{\ZZ_p}} g_S!} w\p{g}
  = \sum_{\substack{a,b,c,d,e\in\ZZ_N\\
      \abs{\set{a,b,c,d,e}}=5}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}.
  \label{eq:diagram_definition}
\end{align}
The proportionality factor accounts for the number of ways to permute
the indices within each region $S\in\PP\p{\ZZ_p}$.  Including this
symmetry factor in the definition of a diagram is a matter of choice
that will ultimately make it easier to work with diagrams.  A diagram
such at that in \eqref{eq:diagram_definition} with $\abs{g}$ dots
nominally takes $\O\p{N^{\abs{g}}}$ time to compute.  In practice,
this computational cost can be prohibitive for performing numerical
simulations of systems of any appreciable size.  If possible, we
therefore wish to simplify diagrams in order to reduce their
computational complexity.  To this end, we first need to define simple
modifications of diagrams in which dots may be replaced by empty dots
(``$\circ$'') or crosses (``$\bm\times$'').

In any diagram (with filled dots, empty dots, or crosses), the values
of indices associated with filled dots are always summed as if the
diagram had only dots, without regard for all other indices.  Empty
dots in a diagram then correspond to indices that are summed over all
of $\ZZ_N$ independently and without additional constraints, e.g.
\begin{align}
  \diagram{example_triple_oo_2} \equiv
  \sum_{a,c\in\ZZ_N}
  \sum_{\substack{b,d,e\in\ZZ_N\\\abs{\set{b,d,e}}=3}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}.
\end{align}
Crosses, meanwhile, correspond to indices that are {\it maximally
  dependent} on filled dots: ``cross indices'' are summed {\it only}
over the values of ``filled dot indices'', e.g.
\begin{align}
  \diagram{example_triple_xx_1}
  \equiv \sum_{\substack{a,d,e\in\ZZ_N\\\abs{\set{a,d,e}}=3}}
  \sum_{b,c\in\set{a,d,e}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e}.
\end{align}
Empty dots and crosses allow us to decompose diagrams into sums of
different diagrams that (may) require less time to evaluate.  We will
essentially use two tricks to decompose diagrams: eliminating filled
dots, and eliminating crosses.

Given a diagram containing only dots (filled or empty), we can
decompose any filled dot into an empty dot and a cross.  For example
\begin{align}
  \diagram{example_triple}
  = \diagram{example_triple_o_top} - \diagram{example_triple_x_top},
  \label{eq:example_decomposition}
\end{align}
where
\begin{align}
  \diagram{example_triple}
  \equiv \sum_{\substack{a,b,c,d,e\in\ZZ_N\\\abs{\set{a,b,c,d,e}}=5}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
\end{align}
\begin{align}
  \diagram{example_triple_x_top}
  \equiv \sum_{\substack{b,c,d,e\in\ZZ_N\\\abs{\set{b,c,d,e}}=4}}
  \sum_{a\in\set{b,c,d,e}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
\end{align}
\begin{align}
  \diagram{example_triple_o_top}
  \equiv \sum_{\substack{b,c,d,e\in\ZZ_N\\\abs{\set{b,c,d,e}}=4}}
  w_1\p{\circ,b,c} w_2\p{b,d,e} w_3\p{b,c,d,e},
  \label{eq:example_triple_x_top}
\end{align}
with
\begin{align}
  w_1\p{\circ,b,c} \equiv \sum_{a\in\ZZ_N} w_1\p{a,b,c},
\end{align}
The decomposition in \eqref{eq:example_decomposition} essentially
breaks up a sum over $a\in\ZZ_N\setminus\set{b,c,d,e}$ into a
difference of sums over $a\in\ZZ_N$ and $a\in\set{b,c,d,e}$.  The sum
over $a\in\ZZ_N$ to compute $w_1\p{\circ,b,c}$ has $\O\p{N^3}$ cost,
and can be performed prior to computing the diagram in
\eqref{eq:example_triple_x_top}.  The decomposition in
\eqref{eq:example_decomposition} thereby reduces the calculation of
one $\O\p{N^5}$ diagram to the calculation of two $\O\p{N^4}$
diagrams.  A similar reduction by a factor of $\sim N$ can be achieved
by decomposing any dot in a region $S\in\PP\p{\ZZ_p}$ addressing only
one tensor, i.e.~with $\abs{S}=1$.

While it is possible, in principle, to derive rules for decomposing
filled dots in a diagram that contains crosses, it is much easier to
simply eliminate crosses from a diagram altogether.  To eliminate a
cross from a diagram, we first note that the corresponding index need
not sum over the values of indices in any tensor that contains that
index; that is (hint: pay attention to the second sum on each line),
\begin{align}
  \diagram{example_triple_x_top_elim}
  &\equiv \sum_{\substack{b,c,d,e,f\in\ZZ_N\\\abs{\set{b,c,d,e,f}}=5}}
  \sum_{a\in\set{b,c,d,e,f}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,f} \\
  &= \sum_{\substack{b,c,d,e,f\in\ZZ_N\\\abs{\set{b,c,d,e,f}}=5}}
  \sum_{a\in\set{d,e,f}}
  w_1\p{a,b,c} w_2\p{b,d,e} w_3\p{b,c,f}.
  \label{eq:x_elim}
\end{align}
For each remaining index addressed by a cross, meanwhile, the cross
simply forces the corresponding indices to be equal, so
\begin{align}
  \diagram{example_triple_x_top_elim}
  = 2 \diagram{example_triple_x_lft_elim}
  + \diagram{example_triple_x_rht_elim},
\end{align}
where the factor of 2 appears simply to account for both of
$a\in\set{d,e}$ in \eqref{eq:x_elim}.  The rules derived here for
replacing filled dots and crosses will be used in the following
sections to simplify diagrams and minimize the computational cost of
their evaluation.

@@

Calculating the second-order effective Hamiltonian $H_1^{(2)}$ induced
on the permutationally symmetric manifold $\M_0$ by the perturbation
$\V_1$ in \eqref{eq:perturbations} requires us to simplify a product
of the form
\begin{align}
  \P_0 \sp{\sum_{p,q} v_p w_q X_p Y_q} \P_0
  \EQFS \diagram{single_body_2} X_1 Y_2
  + \diagram{single_body_1} X_1 Y_1,
  \label{eq:PXYP_start}
\end{align}
where $p,q$ index individual spins; $X,Y$ are single-spin operators;
and the coefficients $v_p,w_q$ satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta\ne0$, which implies that they are mean-zero.
Defining $\v m \equiv \sum_p m_p \ket{p}$ for $m\in\set{v,w}$, we can
simplify
\begin{align}
  \diagram{single_body_1}
  \equiv \sum_p v_p w_p
  = \v v\c\v w,
\end{align}
and
\begin{align}
  \diagram{single_body_2}
  = \diagram{single_body_2_o} - \diagram{single_body_2_x}
  = - \diagram{single_body_1}
  = - \v v \c\v w,
\end{align}
so
\begin{align}
  \sum_{p,q} v_p w_q X_p Y_q
  \EQFS - \v v\c\v w \p{X_1 Y_2 - X_1 Y_1},
\end{align}
In order to write this result in terms of collective operators
$\col{Z} \equiv \sum_p Z_p$, we expand
\begin{align}
  \col{X Y} \EQFS N X_1 Y_1,
  &&
  \col{X}\,\col{Y}
  = \sum_{p=q} X_p Y_q + \sum_{p\ne q} X_p Y_q
  \EQFS \col{XY} + N\p{N-1} X_1 Y_2,
\end{align}
which implies that
\begin{align}
  \sum_{p,q} v_p w_q X_p Y_q
  \EQFS - \f{\v v \c\v w}{N\p{N-1}}
  \p{\col{X}\,\col{Y} - N \col{XY}}.
\end{align}

@@

Calculating the second-order effective Hamiltonian $H_2^{(2)}$ induced
on the permutationally symmetric manifold $\M_0$ by the perturbation
$\V_2$ in \eqref{eq:perturbations} requires us to simplify a product
of the form
\begin{multline}
  \P_0 \sp{\sum_{k<\ell} \sum_{p<q}
    v_{k\ell} w_{pq} O_{k\ell} Q_{pq}} \P_0 \\
  \EQFS \f14 \diagram{two_body_4} O_{1,2} Q_{3,4}
  + \diagram{two_body_3} O_{1,2} Q_{1,3}
  + \f12 \diagram{two_body_2} O_{1,2} Q_{1,2},
  \label{eq:POQP_start}
\end{multline}
where $k,\ell,p,q$ index individual spins; $O$ and $Q$ are
permutationally-symmetric two-spin operators; the coefficients
$v_{k\ell},w_{pq}$ satisfy \eqref{eq:cond_2} with eigenvalue
$\Delta\ne0$, which implies that they are mean-zero; and the factors
of $1/4$ and $1/2$ are included to cancel out the symmetry factors
that appear in the definition of a diagram.  Defining
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_p m_p \ket{p},
  &&
  m_p \equiv \sum_q m_{pq},
\end{align}
for $m\in\set{v,w}$, we can write
\begin{align}
  \diagram{two_body_2} = \sum_{p,q\in\ZZ_N} v_{pq} w_{pq}
  = 2 \sum_{p,q\in\C_N\p{2}} v_{pq} w_{pq}
  = \v{\m v} \c\v{\m w},
\end{align}
and simplify
\begin{align}
  \diagram{two_body_3}
  &= \diagram{two_body_3_o} - \diagram{two_body_3_x} \\
  &= \diagram{two_body_3_oo} - \diagram{two_body_3_ox}
  - \diagram{two_body_2} \label{eq:mid_zero} \\
  &= \v v \c \v w - 2 \v{\m v} \c \v{\m w},
\end{align}
where the middle diagram in \eqref{eq:mid_zero} contains an empty sum
for the cross, and is therefore equal to zero.  Similarly neglecting
diagrams with empty sums and additionally using the fact that
$v_{pq},w_{pq}$ are mean-zero, we can work out
\begin{align}
  \diagram{two_body_4}
  &= \diagram{two_body_4_o} - \diagram{two_body_4_x} \\
  &= \sp{\diagram{two_body_4_oo} - \diagram{two_body_4_ox}}
  - 2 \sp{\diagram{two_body_3}} \\
  &= \sp{-2 \diagram{two_body_3_o}}
  - 2 \sp{\diagram{two_body_3_oo} - \diagram{two_body_2}} \\
  &= -4 \diagram{two_body_3_oo} + 2 \diagram{two_body_2} \\
  &= -4 \v v\c\v w + 4 \v{\m w}\c\v{\m w}.
\end{align}
Altogether, we thus have that
\begin{multline}
  \sum_{k<\ell} \sum_{p<q} v_{k\ell} w_{pq} O_{k\ell} Q_{pq} \\
  \EQFS - \p{\v v \c \v w - \v{\m v} \c \v{\m w}} O_{1,2} Q_{3,4}
  + \p{\v v\c\v w - 2 \v{\m v} \c \v{\m w}} O_{1,2} Q_{1,3}
  + \v{\m v} \c \v{\m w}\, O_{1,2} Q_{1,2}.
  \label{eq:POQP}
\end{multline}
